---
title: "Homogenise the format of summary statistics from GWAS with *hssGWAS*"
author: "Alan Murphy and Nathan Skene"
date: "`r Sys.Date()`"
bibliography: inst/extdata/hssGWAS.bib
csl: inst/extdata/nature.csl
output: rmarkdown::github_document
vignette: >
  %\VignetteIndexEntry{hssGWAS}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}    
---

<!-- Readme.md is generated from Readme.Rmd. Please edit that file -->

```{r, echo=FALSE}
knitr::opts_chunk$set(tidy         = FALSE,
                      warning      = FALSE,
                      message      = FALSE)
```

# Introduction

The *hssGWAS* package is designed to facilitate the standardisation of GWAS summary statistics as utilised in our Nature Genetics paper [@Skene2018].

# Overview

The package is designed to handle the lack of standardisation of output files by the GWAS community. There is a group who have now manually standardised many GWAS: [R interface to the IEU GWAS database API • ieugwasr](https://mrcieu.github.io/ieugwasr/) and [gwasvcf](https://github.com/MRCIEU/gwasvcf) but because a lot of GWAS remain closed access, these repositories are not all encompassing. 

*hssGWAS* provides a framework standardise the format fo any GWAS summary statistics enabling downstream integration and analysis. The package works by addressing the most common discrepancies across summary statistics.

# Installing hssGWAS

The *hssGWAS* package is available from github. To be able to install the package one needs to run the following lines of code:

```
if (!require("devtools")) {
  install.packages("devtools")
}
devtools::install_github("neurogenomics/hssGWAS")
```

You can then load the package and data package:

```{r }
library(hssGWAS)
```


# Getting started

See the vignette for use cases of *hssGWAS*.

If you have any problems please do file an issue here on github.

# Citation

If you use the hssGWAS package as well then please cite

[Skene, et al. Genetic identification of brain cell types underlying schizophrenia.
Nature Genetics, 2018.](https://www.nature.com/articles/s41588-018-0129-5)


# References
