---
title: "Standardise the format of summary statistics from GWAS with *MungeSumstats*"
author: "Alan Murphy and Nathan Skene"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
csl: inst/cit/nature.csl
vignette: >
    %\VignetteIndexEntry{Standardise the format of summary statistics from GWAS with MungeSumstats} 
    %\usepackage[utf8]{inputenc}
    %\VignetteEngine{knitr::knitr} 
bibliography: inst/cit/MungeSumstats.bib
editor_options: 
  markdown: 
    wrap: 72
---

```{r style, echo=FALSE, results='asis', message=FALSE}
BiocStyle::markdown()
knitr::opts_chunk$set(tidy         = FALSE,
                      warning      = FALSE,
                      message      = FALSE)

```

# Citation

If you use the MungeSumstats package, please cite

[Skene, et al. Genetic identification of brain cell types underlying
schizophrenia, 2016.](https://www.nature.com/articles/s41588-018-0129-5)

# Overview

The *MungeSumstats* package is designed to facilitate the
standardisation of GWAS summary statistics as utilised in our Nature
Genetics paper [@Skene2018].

The package is designed to handle the lack of standardisation of output
files by the GWAS community. There is a group who have now manually
standardised many GWAS: [R interface to the IEU GWAS database API •
ieugwasr](https://mrcieu.github.io/ieugwasr/) and
[gwasvcf](https://github.com/MRCIEU/gwasvcf) but because a lot of GWAS
remain closed access, these repositories are not all encompassing.

The [GWAS-Download
project](https://github.com/mikegloudemans/gwas-download) has collated
summary statistics from 200+ GWAS. This repository has been utilsed to
identify the most common formats, all of which can be standardised with
*MungeSumstats*.

Moreover, there is an emerging standard of VCF format for summary
statistics files with multiple, useful, associated R packages such as
*vcfR*. However, there is currently no method to convert VCF formats to
a standardised format that matches older approaches.

The *MungeSumstats* package standardises both VCF and the most common
summary statistic file formats to enable downstream integration and
analysis.

# Aim

*MungeSumstats* will ensure that the necessary columns are present,
syntactically correct. The output columns will depend on your input but can 
include: 
  * SNP : SNP ID (rs IDs) 
  * CHR : Chromosome number 
  * BP : Base pair positions
  * A1 : Effect allele 
  * A2 : Non-effect allele
  * Z : Z-score
  * N : Sample size
  * P : Unadjusted p-value for SNP
  * BETA : Effect size estimate relative to the alternative allele 

Tests run by *MungeSumstats* include:
  * Check VCF format
  * Check tab delimited
  * Check header names
  * Check for multiple models or traits in GWAS
  * Check for CHR:BP:A2:A1 all in one column
  * Check for CHR:BP in one column
  * Check if CHR and/or BP is missing (infer from SNP ID)
  * Check if SNP ID is missing (infer from CHR and BP)
  * Check that vital columns are present (SNP,CHR,BP,P,A1,A2)
  * Check for one signed column (Z,OR,BETA,LOG_ODDS,SIGNED_SUMSTAT)
  * Check column ordering
  * Check for missing data
  * Check for duplicated columns
  * Check for small p-values (lower than  5e-324)
  * Check N column is an integer
  * Check SNPs are RS ID's
  * Check for duplicated rows, based on SNP ID

If a test is failed, the user will be notified and if possible, the input will 
be corrected.

# Data

The *MungeSumstats* package contains a small subset of the GWAS summary
statistics file on Educational Attainment by Okbay et al 2016: PMID:
27898078 PMCID: PMC5509058 DOI: 10.1038/ng1216-1587b.

This data will be used to showcase MungeSumstats functionality.

# Running MungeSumstats

The R package can be installed from github as follows:

    if (!require("devtools")) {
      install.packages("devtools")
    }
    devtools::install_github("neurogenomics/MungeSumstats")

```{r setup}
library(MungeSumstats)
#save the Educational Attainment Okbay sumstat file to a temp directory
tmp <- tempfile()
writeLines(MungeSumstats::eduAttainOkbay,con = tmp)
```

To standardise the summary statistics' file format, simply call
`format_sumstats()` passing in the path to your summary statistics file.
The path to the reformatted summary statistics file will be returned

```{r, message=TRUE}
reformatted <- MungeSumstats::format_sumstats(tmp)
```

# Future Enhancements

The *MungeSumstats* package aims to be able to handle the most common
summary statistic file formats including VCF. If your file can not be
formatted by *MungeSumstats* feel free to report the bug on github:
<https://github.com/neurogenomics/MungeSumstats>

# References
