---
title: "Standardise the format of summary statistics from GWAS with *MungeSumstats*"
author: "Alan Murphy and Nathan Skene"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
csl: inst/cit/nature.csl
vignette: >
    %\VignetteIndexEntry{Standardise the format of summary statistics from GWAS with MungeSumstats} 
    %\usepackage[utf8]{inputenc}
    %\VignetteEngine{knitr::knitr} 
bibliography: inst/cit/MungeSumstats.bib
editor_options: 
  markdown: 
    wrap: 72
---

```{r style, echo=FALSE, results='asis', message=FALSE}
BiocStyle::markdown()
knitr::opts_chunk$set(tidy         = FALSE,
                      warning      = FALSE,
                      message      = FALSE)

```

# Citation

If you use the MungeSumstats package, please cite

[Skene, et al. Genetic identification of brain cell types underlying
schizophrenia, 2016.](https://www.nature.com/articles/s41588-018-0129-5)

# Overview

The *MungeSumstats* package is designed to facilitate the
standardisation of GWAS summary statistics as utilised in our Nature
Genetics paper [@Skene2018].

The package is designed to handle the lack of standardisation of output
files by the GWAS community. There is a group who have now manually
standardised many GWAS: [R interface to the IEU GWAS database API •
ieugwasr](https://mrcieu.github.io/ieugwasr/) and
[gwasvcf](https://github.com/MRCIEU/gwasvcf) but because a lot of GWAS
remain closed access, these repositories are not all encompassing.

The [GWAS-Download
project](https://github.com/mikegloudemans/gwas-download) has collated
summary statistics from 200+ GWAS. This repository has been utilsed to
identify the most common formats, all of which can be standardised with
*MungeSumstats*.

Moreover, there is an emerging standard of VCF format for summary
statistics files with multiple, useful, associated R packages such as
*vcfR*. However, there is currently no method to convert VCF formats to
a standardised format that matches older approaches.

The *MungeSumstats* package standardises both VCF and the most common
summary statistic file formats to enable downstream integration and
analysis.

*MungeSumstats* also offers a number of Quality Control (QC) steps 
which are important prerequisites for downstream analysis like Linkage 
disequilibrium score regression (LDSC) and MAGMA.

# Aim

*MungeSumstats* will ensure that the all essential columns for analysis are 
present and syntactically correct. Generally, summary statistic files include 
(but are not limited to) the columns: 
  * SNP : SNP ID (rs IDs) 
  * CHR : Chromosome number 
  * BP : Base pair positions
  * A1 : Effect allele 
  * A2 : Non-effect allele
  * Z : Z-score
  * BETA : Effect size estimate relative to the alternative allele 
  * P : Unadjusted p-value for SNP
  * N : Sample size
  * INFO: The imputation information score
  * FRQ: The minor allele frequency (MAF) of the SNP

Tests run by *MungeSumstats* include:
  * Check VCF format
  * Check tab or space delimited
  * Check header names
  * Check for multiple models or traits in GWAS
  * Check for CHR:BP:A2:A1 all in one column
  * Check for CHR:BP in one column
  * Check if CHR and/or BP is missing (infer from reference genome)
  * Check if SNP ID is missing (infer from reference genome)
  * Check if A1 and/or A2 are missing (infer from reference genome)
  * Check that vital columns are present (SNP,CHR,BP,P,A1,A2)
  * Check for one signed/effect column (Z,OR,BETA,LOG_ODDS,SIGNED_SUMSTAT)
  * Check for missing data
  * Check for duplicated columns
  * Check for small p-values (lower than  5e-324)
  * Check N column is an integer
  * Check SNPs are RS ID's
  * Check for duplicated rows, based on SNP ID
  * Check for duplicated rows, based on base-pair position
  * Check for SNPs on reference genome
  * Check INFO score
  * Check for strand-ambiguous SNPs
  * Check for non-biallelic SNPs (infer from reference genome)
  * Check for allele flipping
  * Check for SNPs on chromosome X, Y, and mitichondrial SNPs (MT)
  * Check for SNPs with  
  
If a test is failed, the user will be notified and if possible, the input will 
be corrected. The QC steps from the checks above can also be adjusted to suit 
the user's analysis, see `MungeSumstats::format_sumstats`.

# Data

The *MungeSumstats* package contains small subsets of GWAS summary
statistics files. Firstly, on Educational Attainment by Okbay et al 2016: PMID:
27898078 PMCID: PMC5509058 DOI: 10.1038/ng1216-1587b.

Secondly, a VCF file (VCFv4.2)relating to the GWAS Amyotrophic lateral 
sclerosis from ieu open GWAS project. Dataset: ebi-a-GCST005647:
https://gwas.mrcieu.ac.uk/datasets/ebi-a-GCST005647/

These datasets will be used to showcase MungeSumstats functionality.

# Running MungeSumstats

The *MungeSumstats* is in the process of being added to Bioconductor but, in the 
meantime, is available from github. To be able to install the package one needs 
to install the devel version of R (version 4.1) which can be found at 
https://cran.r-project.org/ and then run the following lines of code:

```
if (!require("devtools")) {
  install.packages("devtools")
}
devtools::install_github("neurogenomics/MungeSumstats")
```

To install *MungeSumstats* on Bioconductor run:

```
if (!require("BiocManager"))
    install.packages("BiocManager")

BiocManager::install(version = "devel")
BiocManager::install("MungeSumstats")
```
Once installed, load the package and the sample dataset, writing to a temporary directory:

```{r setup}
library(MungeSumstats)
#save the Educational Attainment Okbay sumstat file to a temp directory
tmp <- tempfile()
writeLines(MungeSumstats::eduAttainOkbay,con = tmp)
```

To standardise the summary statistics' file format, simply call
`format_sumstats()` passing in the path to your summary statistics file.
You must also specify which genome build was used in the GWAS(GRCh37 or GRCh38).
The reference genome is used for multiple checks like deriving missing 
data such SNP/BP/CHR/A1/A2 and for QC steps like removing non-biallelic
SNPs or strand-ambiguous SNPs. The path to the reformatted summary statistics 
file is returned by the function call.

```{r, message=TRUE}
reformatted <- MungeSumstats::format_sumstats(path=tmp,ref_genome="GRCh37")
```

The hyperparameters `format_sumstats` in that control the level of QC conducted 
by *MungeSumstats* are:
  * *convert_small_p* Binary, should p-values < 5e-324 be converted to 0? Small p-values pass the R limit and can cause errors with LDSC/MAGMA and should be converted. Default is TRUE.
  * *convert_n_int* Binary, if N (the number of samples) is not an integer, should this be rounded? Default is TRUE.
analysis_trait If multiple traits were studied, name of the trait for analysis from the GWAS. Default is NULL
INFO_filter numeric The minimum value permissible of the imputation information score (if present in sumstatsfile). Default 0.9
  * *N_std* Numeric, the number of standard deviations above the mean a SNP's N is needed to be removed. Default is 5.
rmv_chr vector or character The chromosomes on which the SNPs should be removed. Use NULL if no filtering necessary. Default is X, Y and mitochondrial. 
  * *on_ref_genome* Binary, should a check take place that all SNPs are on the reference genome by SNP ID. Default is TRUE
  * *strand_ambig_filter* Binary, should SNPs with strand-ambiguous alleles be removed. Default is FALSE
  * *allele_flip_check* Binary, should the allele columns be chacked against reference genome to infer if flipping is necessary. Default is TRUE
  * *bi_allelic_filter* Binary, should non-biallelic SNPs be removed. Default is TRUE

VCF files can also be standardised to the same format as other summary statistic
files. A subset of the Amyotrophic lateral sclerosis GWAS from the ieu open GWAS
project has been added to *MungeSumstats* to demonstrate this functionality.
Simply pass the path to the file in the same manner you would for other summary
statistic files:

```{r, message=TRUE}
#save ALS GWAS from the ieu open GWAS project to a temp directory
tmp2 <- tempfile()
writeLines(MungeSumstats::ieuAmlVcf,con = tmp2)
reformatted_vcf <- MungeSumstats::format_sumstats(path=tmp2,ref_genome="GRCh37")
```

# Future Enhancements

The *MungeSumstats* package aims to be able to handle the most common
summary statistic file formats including VCF. If your file can not be
formatted by *MungeSumstats* feel free to report the bug on github:
<https://github.com/neurogenomics/MungeSumstats> along with your summary
statistic file header. 

We also encourage people to edit the code to resolve their particular issues 
too and are happy to incorporate these through pull requests on github. If your
summary statistic file headers are not recognised by *MungeSumstats* but 
correspond to one of SNP, BP, CHR, A1, A2, P, Z, OR, BETA, LOG_ODDS,
SIGNED_SUMSTAT, N, N_CAS, N_CON, NSTUDY, INFO or FRQ, feel free to update the 
`MungeSumstats::sumstatsColHeaders` following the approach in the data.R file
and add your mapping. Then use a pull request on github and we will incorporate 
this change into the package.



# References
